Efficient pedigree recording for fast population genetics simulation

### 2017-11-16 forwards_paper.tex
added space complexity figure

Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals 
using Algorithm~\algref{W}
for $T$ generations,
with the probability of death, $\delta$, set to 1 for simplicity.
Since there is exactly one crossing over every generation,
this produces tables of 
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
so that $T \sim 20N$, say.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
This tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
three edges end and three new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
Coalescent theory tells us that 
the expected total length of edges in a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per generation.
Not all such recombinations actually change the tree topology,
and three times this gives us an upper bound on the expected number of edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $160 N^2$ items to $6N(4\log(2N) + 3)$;
at $N=10^4$, a factor of 3,000.

>>>>

Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals 
using Algorithm~\algref{W}
for $T$ generations,
with the probability of death, $\delta$, set to 1 for simplicity.
Since there is exactly one crossing over every generation,
this produces tables of 
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
so that $T \sim 20N$, say.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
This tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
three edges end and three new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
Coalescent theory tells us that 
the expected total length of edges in a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per generation.
Not all such recombinations actually change the tree topology,
and three times this gives us an upper bound on the expected number of edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $160 N^2$ items to $8N(4\log(2N) + 3)$;
at $N=10^4$, a factor of 4,700.

To get an idea of how required space depends on the length of time,
suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
This produces $4NT$ edges; how many are required after simplification?
As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
Roughly, the length of time for which a marginal tree has k tips is 
$2N/k(k-1) = 2N(1/(k-1) - 1/k)$,
and so the time to go from N to n tips is $2N(1/(n-1) - 1/(N-1))$.
This implies that after running for time $T$ we expect to 
have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is 
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which 
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align*}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align*}
This holds up quite well in practice --
the number of edges actually required for 50 independent Wright--Fisher simulations,
as a function of time (obtained by running simplify every generation)
is shown in \ref{fig:num_edges}, compared to this prediction.

\begin{figure}
    \begin{center}
        \includegraphics{sims/num_edges_50}
    \end{center}
    \caption{
        Number of edges in the simplified tree sequence
        for 50 Wright-Fisher simulations begun with no prior history
        as a function of number of generations.
        Each line is one simulation, the black line gives the average,
        and the dotted line is the upper bound of the text.
        \label{fig:num_edges}
    }
\end{figure}
